Lossy compression of DNA sequencing quality data

### 2015-02-04 manuscript/q-compression.tex
Integrated the economic costs of memory into the subsections for each category of compression we examine.

Quality values are typically used throughout bioinformatics pipelines.
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie 2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

@@

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two, i.e., two
categories ``good'' and ``bad'' quality. We implement 2-bin encoding
by setting a quality value threshold empirically determined by the
distribution of quality values across reads. Base-calls are marked
``bad'' if their quality value falls below the first quartile minus
1.5 $\times$ the interquartile range (IQR), which is the difference
between the first and third quartile. 1.5 $\times$ IQR is the value
used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
of this approach is that it is completely data-dependent. No prior
assumptions of the distribution of the quality values need to be made.

Because we only have two possible choices for a quality of base, this
makes a straightforward binary encoding possible, allowing us to use a
single bit to represent the quality of a base instead of the default 8
bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.

Although simple and intuitive, considering a base as either good or
bad may be too limiting for some users, so we provide a few additional
options to preserve more of the defining characteristics of the
quality values, but we only analyze 2-bin encoding in our results.

@@

@@

Although this approach is not able to capture randomly occuring
outlier quality values, it ensures that the overall trends in quality
sequences are retained. Quality profiles capture different overall
qualities and different drop-off positions and gradients. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

@@

\subsection{The economics of lossy quality sequence compression}

From an economics perspective, we can examine the fixed and marginal
costs associated with our lossy compression techniques with respect to
memory and computation. For our quality value binning and model-based
compression methods there are no explicit fixed costs in terms of
memory, unlike the profile--based approach, which has a small fixed
cost in storing the profiles. However, the profile-based approach has
a much smaller marginal compared to the rest due to the fact that we
are simply storing the profile index for each quality sequence.

Regarding computational costs, only quality binning and the
profile--based methods have associated fixed costs. Before performing
compression, we have to determine the optimal thresholding values for
quality binning, and the cluster representatives for the
profile--based method.

In practice, the specific values of these costs are very sensitive to
the implementation details.

>>>>

Quality values are typically used throughout bioinformatics pipelines.
Among the most fundamental uses of sequence quality values is as part
of the quality assessment and quality control (\textsc{qa/qc})
processes prior to subsequent analysis steps. Quality control based on
quality values generally includes two operations:
\textit{i}.~filtering, the elimination of reads that on the whole do
not meet quality standards, which reduces the total number of reads;
and \textit{ii}.~trimming of low quality base-calls from reads, which
reduces the number total number of bases. Quality values can be used
by genome assembly software to produce better
assemblies~\cite[e.g.,][]{Bryant:2009uq,Gnerre:2011kx}. Short-read
alignment software, such as Bowtie2~\cite{Langmead:2012rw}, use
quality values to weight mismatches between query and reference
sequences.  Software for detecting single nucleotide polymorphisms
(\textsc{snp}s) can use quality values, and identified \textsc{snp}s
with high-quality calls are deemed more reliable than those with
low-quality calls, particularly in low coverage
regions~\cite[e.g.,][]{McKenna:2010bh}.

@@

Quality values can be binned, and the minimum number of bins that
allows for a any distinction among quality values is two, i.e., two
categories ``good'' and ``bad'' quality. We implement 2-bin encoding
by setting a quality value threshold empirically determined by the
distribution of quality values across reads. Base-calls are marked
``bad'' if their quality value falls below the first quartile minus
1.5 $\times$ the interquartile range (IQR), which is the difference
between the first and third quartile. 1.5 $\times$ IQR is the value
used by Tukey's box plot~\cite{mcgill1978variations}. The main benefit
of this approach is that it is completely data-dependent, and no
assumptions regarding the distribution of the quality values need to be made.

With 2-bin encoding binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard 8
bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,DBLP:journals/tit/ZivL78}.

The economic costs of memory use for binning, in general terms,
include no fixed costs, and marginal costs that are a function of the
number of base-call quality values encoded times the cost of the
encoding.

@@

The economic costs of memory use for model-based compression, in
general terms, include no fixed costs, and marginal costs that are a
function of the number of reads times the cost representing the model
parameters.

@@

Although this approach is not able to capture randomly occurring
outlier quality values, it ensures that the overall trends in quality
sequences are retained. Quality profiles capture different overall
qualities and different drop-off positions and gradients. An example
of 128 quality profiles are shown on Figure \ref{fig:profiles_128}.

The economic costs of memory use for profile-based compression, in
general terms, include fixed costs associated with representing the
profiles, which is a function of the number of profiles times the cost
of encoding them, and these fixed costs are amortized over the entire
set of reads to which they apply. Additionally there are marginal
costs that are a function of the number of reads encoded.

@@
